Preregistration for modelling in ecolgy and conservation

### 2023-09-27 _analysis/index.Rmd
rename _analysis

---
title: "Preregistration for modelling in ecolgy and conservation"
description: |
  Working Draft.
author:
  - name: Elliot Gould 
    affiliation: School of BioSciences, University of Melbourne
  - name: Chris Jones
    affiliation: Department of Environment Land and Water
  - name: Libby Rumpff
    affiliation: School of BioSciences, University of Melbourne
  - name: Hannah Fraser
    affiliation: School of BioSciences, University of Melbourne
  - name: Fiona Fidler
    affiliation: School of BioSciences, University of Melbourne
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    self_contained: false
---
**Intended journal:** Methods in Ecology and Evolution.

@@

We should be concerned about research transparency because, a lack of complete and transparent reporting practices provide the conditions in which QRPs are likely to arise [@Fidler:2017he]. Moreover, as Morrison [-@Morrison:2016cd] demonstrated, when models are not reported transparently with sufficient detail, they cannot be reproduced. This is particularly problematic because in ecology, models are typically used in decision-support, often directly underpinning policy and management decisions. Incomplete and non-transparent reporting of the model and the modelling process risk the model being used in inappropriate applications or decision contexts [@Grimm:2014es] because the underlying modelling choices and assumptions are not communicated [@deVos:2011]. Alternatively, a lack of transparency may mask poor model design or serious flaws in the model, with the risk that they be used in decision support resulting in wrong, or irreversible decisions [@Grimm:2014es;@Schmolke:2010fd]. A lack of transparency in reporting also risks undermining the credibility of the model, such that the decision-maker rejects the use of the model and instead turns to unstructured and subjective judgments [@Grimm:2014es;@Addison:2013wa], thus contributing further to "decision complacency", or the "knowing-doing" gap [@Gardner:2018dm].

Preregistration is a practice that has been adopted in disciplines that have confronted the 'reproducibility crisis' head-on, and can improve the transparency and reproducibility of ecological models. Preregistration requires the methods and analysis plan of a future study are registered in a secure and accessible platform prior to any data collection, and after submission it can no longer be altered. Preregistration distinguishes between what is a genuine a priori planned analysis, and what is not (Parker, Fraser, and Nakagawa 2019), and therefore prevents the opportunistic misuse of researcher degrees of freedom or 'Questionable Research Practices' (QRPs, Wicherts et al., 2016). Researcher degrees of freedom refer to the many analytic decisions a researcher makes during the planning, data collection, data analysis and reporting of results in scientific studies (Wicherts et al. 2016; Simmons, Nelson, and Simonsohn 2011). Examples of these analytic choices include whether to exclude observations, whether more data should be collected, which set of conditions should be combined and compared, which model evaluation metric should be used, etc. Coupled with a lack of transparency -- or "undisclosed flexibility" -- researcher degrees of freedom are vulnerable to being exploited by researchers because of cognitive biases and incentive structures within science. \<lead in sentence to next para??\>

Ecology has seen recent improvements in journal policy concerning reporting guidelines and data/code archiving, but there has been slow uptake of other important tools, like preregistration and registered reports (Parker, Fraser, and Nakagawa 2019). Pu et al. 2019 found that conceptual barriers may prevent users from different disciplines to think that preregistration is "not for them", this was a particularly salient sentiment for researchers engaged in non-experimental studies. We have experienced this sentiment ourselves when looking at preregistration templates to use for model-based research --- nearly all preregistration templates are geared towards experimental hypothesis-testing research. Such templates contain items that are simply irrelevant to model-based and/or applied ecological research. More importantly, they do not reflect the iterative nature of model development [@MacEachern2019] and fail to capture critical analytic decisions in the modelling process. Presently, we are aware of only a single preregistration template for modelling [@Cruwell2019b], however it is steeped heavily in an an academic psychology context, is focussed on the use of models for hypothesis-testing, neglects the entire model development process, and approaches modelling from a very narrow epistemic function, in a way that is not commonly applied in ecology.

Preregistration cannot simply be ported over to other disciplines and adopted as-is. If we wish to adopt preregistration in ecology, specifically within model-based research, both the content required to be preregistered, as well as the procedure for creating a preregistration must be adapted to suit model-based research as it is undertaken in ecology. Different research methodologies, and their specific and nuanced application across different domains and fields of study follow different analytical choices and are therefore susceptible to different QRPs. Consequently, domain-specific templates are necessary for guiding preregistration in such a way that restricts researcher degrees of freedom as they uniquely emerge within a domain or field-specific application of a particular research methodology.

@@

The goal of this study is to \<rethink\> preregistration to better suit the domain-specific challenges and methodological particularities of model-based research in ecology. This research builds on recently developed "QRP roadmaps" (Gould et al. 2019) and research by Cruwell et al. (2019) showing that existing templates are too general, and do not account for domain-specific and research category-specific nuances and questionable research practices. We aim to develop a template for preregistering ecological models, and devise a methodology for implementing an expanded view of the preregistration process, 'adaptive preregistration' (*sensu* Srivastava), that better accounts for the inherently iterative nature of model development than static, deterministic document-based preregistration.

@@

The secondary purpose of the model is to increase *research transparency* of model-based research in ecology and conservation. *Research transparency* refers to both production transparency --- which includes research artefacts like open-access data and materials, or data collection procedures --- and analytic transparency --- which describes a complete account of how the analytic conclusions are drawn from the data (Lupia and Elman, 2014). \<refs to model reporting literature- it's dire, can't properly be evaluated by a reviewer, moreover can't be reproduced\>

@@

More specifically, we intend this template to be suitable not just in 'pure-research'-only contexts, but to be suitable for ecologists and conservation decision-makers who are working in the 'actionable sciences' \<which guy was this again.. Sutherland?? define\>.  The template should be adequate for use by scientists and modellers working in applied ecological contexts, but also for management and conservation decision-making contest where there are real-world decisions hinging on the model-based research.

\<Research Types: Modelling - what do we mean by modelling?? I think if we don't specify this then people will just think 'classical linear regression' as modelling, and then will say, well why don't use existing hypothesis testing templates, like AsPredicted. I need to specify some particularities about the type of modelling that occurs in in these research contexts, and why existing templates don't suit, and that we need to design specifically for these research contexts.\>

@@

Consequently, we aim not just to simply design a new template, deciding on its structure and content, but we propose a workflow and methodology to be implemented by the author when completing a preregistration of their model-based research \<see section - inner logic of the model - discuss features model-based research, like iterative development?\>

@@

**Identifying aspirations and desirable attribute for the template**

What attributes do we aim for the template

A) an interface that ensures ease of use for the creator of a preregistration

B) structure and content that is necessarily opinionated.. forces user to a particular workflow, so getting that workflow right, that reflects both norms and practice of model-based research is pivotal.

On the 4^th^ March 2020, we ran a series of structured activities designed to inform the development of two preregistration templates, one for field-work and sampling design, the other for modelling and data-analysis. We split the activities into two groups, to better focus discussion for each of these templates. All workshop materials and workshop outputs can be found on the OSF component here \<LINK\>.

@@

The objective of this activity was to identify modelling workflows that capture critical decision-steps in the process of field sampling efforts and modelling exercises that should be included in a preregistration template. By "critical" we mean that without these decision-steps the template would not a) adequately constrain researcher degrees of freedom and b) transparently document the analytic decision-process. The activity consisted of two-parts, first, we individually reflected on our scientific process for a recent or memorable research project, listing the steps we undertook from the beginning to end.  In the second part of this exercise, we came together as a group to collate each of the steps in our personal workflows onto a group workflow template. This template was prepared earlier (\<specific OSF link to the document\>), and consisted of several key 'phases' of the scientific process, with decision-steps grouped under each phase. The development of the draft modelling workflow was derived from previous work by Gould et al. *in prep.,* whilst the field sampling workflow was created specifically for this exercise. We sorted and categorised each decision-step as belonging to a particular workflow phase. Next we grouped similar decision-steps that could be included as a single item on a preregistration template. At the end of this process we reviewed the suggested phases on the workflow to assess whether phases needed further discretisation, aggregation, removal, or amendment.

@@

This second activity aimed to identify critical challenges and roadblocks to the implementation and uptake of preregistration in ecology and conservation, as well as brainstorming solutions to these challenges. We conducted two lots of structured discussions, one for each of the field and modelling template groups. Each discussion was structured using a series of pre-prepared questions and talking points informed by recent debates about preregistration in the literature \<cite / link to OSF, or even discuss??\>. These talking points were chosen with the aim of identifying problems in the content and or structuring of the phases and decision steps that were collectively identified in activity 1, as well as any conceptual or philosophical issues with preregistration in ecology, conservation, and in particular modelling, that might impact on the template structure, content and procedural implementation of preregistration.

@@

**1\. Identify Scientific Workflows for ecological modelling, both real/in-practice and idealised.**

The goal of this analysis task was to generate a complete draft workflow for both fieldwork and modelling and data-analysis that will form the basis of the final preregistration templates. We used a combination of inductive and deductive coding \<is this what it is called\> to analyse the workshop outputs and generate a model scientific workflow for model-based research that would constitute the presumed scientific workflow that would be used to guide the structure of the preregistration template. We collated and coded the workshop outputs using \<atlas.TI, insert software citation and version\>, \<a qualitative coding software\>. All workshop outputs can be found here (\<Link to Workshop Outputs OSF repository\>).

A preliminary idealised scientific workflow for ecological modelling research was synthesised after reviewing the literature on 'good modelling practice' and structured decision-making literature in ecological management. Steps in the workflow were coded and broken down into 'phases' consisting of 'steps' and 'substeps', in line with the ontology we specified for translating a modelling workflow into a preregistration template [section link], \<insert link to appendices or just present the final?\>. Each decision-point in the personal modelling workflows generated from activity one of the collaborative workshop were coded as belonging to one of these phases, steps and substeps. We revised the coding structure \<inductively (is this the word)\>, according to patterns and themes identified across multiple personal workflows. For example, we added the final phase \<name for last phase\>, which aimed to capture the fact that analysts usually present the results of their modelling to clients, decision-makers or other stakeholders that must interpret the evidence and are responsible for making management decisions. In this way, we identified a common or generalised workflow from different types of applied ecological modelling projects that synthesised both \<idealised norms and norms of practice, Pu et al., see other philosophy of sci refs\>. This workflow is used to provide structure and content to the preregistration template.

**2\. Translating Workflows into Preregistration Templates**

We translated the model-based workflow identified in part I of workshop analysis into a text-based preregistration template. Reflecting on existing templates \<link to OSF repository of templates page\>, we identified common elements / units of preregistration templates that we mapped onto the generalised modelling workflow identified in part I of workshop output analysis. Following the features of a good guideline identified by Moallemi et al. [-@Moallemi2019], we provide examples of choices or illustrations to clarify each step of the preregistration, for parts of the modelling process that are semantically, epistemically and/or philosophically contentious or ambiguous \<link to paper describing contentious issues in modelling history\>. We synthesised the following ontology to build our preregistration template:

1\. Decision Phase - each has a "title" and "description" or "definition", steps are grouped under phases. \<Note, that these are being scrapped from the template and are turned into tool-tips\>

2\. Decision Step - each has a "title" and "description" or "definition" choices are grouped into 'steps', which corresponds to an activity or set of related activities in a scientific workflow. \<again, these are being scrapped and turned into tool-tips.\>

3\. Choices ---\> each 'choice' corresponds to a uniquely numbered item on the PRT requiring a response from user. The style should be in the form of a directive or a question. E.g. "explain how you will do xyzzy". Or "What performance measure will you use for assessing 'goodness of fit'?"

**3\. Designing a new methodology for implementing preregistration in model-based research**

@@

\- What should a research workflow look like for a modeller in an applied ecology setting who wishes to preregister their work?

\- How do we accommodate the iterative cycle of model development into the preregistration process?

\- At what stage in a modelling or structured decision-making process should preregistration begin?

@@

\<which other template design papers have included evaluation criteria and rationale, Sophia's?\>

@@

Modellers \<here, here, and here\> have expressed concern about preregistration making modelling a more difficult task than it already is, and hampering the ability to meet project deadlines, especially in decision-problems. Many modellers already struggle with sufficiently documenting models during publication and reporting \<ref, model reporting and transparency review\>, and modellers working in government or other agencies have expressed concern that preregistration will simply add to the list of  'red-tape' expected of them. Given these concerns, and considering the concerns of many with regard to the iterative nature of model development precluding making preregistration impossible or inappropriate, we figured it was of utmost importance that the templates be also evaluated on their ease-of-use and feasibility.

@@

We tested and evaluated the templates on the ability to meet these criteria by using a model-based research case-study \<see box\>.  Rather than 'digging back' through the version history of a completed preregistration to ask researchers about the process and how it had been developed in detail \<like Pu et al\>, we chose to leverage the version-control and collaborative project management features of Git and GitHub [@BenBondLamberty:2018by; @Blischak:2016kr] as a tool for documenting both the analytic decisions of the researcher, and for using the preregistration process itself to 'live-develop' the template template. For example, if it turned out that we had missed an important item, or perhaps the phrasing of an item or the order and structure of the template needed to change whilst in the process of completing the preregistration, the suggested change, and its justification would be both recorded in GitHub and linked to the actual preregistration. Since *Pu et al* (p16). Found that the process of creating a preregistration is 'difficult to reconstruct with interview questions alone', we we aimed to capture the analysts 'thinking in the moment' by using comments and GitHub issues as to why a particular decision about the template item inclusion and specification should change. \<How we did this using GitHub:How to present this material? Link to the wiki issue and instructions to users?\>

@@

\<TEXT HERE DESCRIBING THE ENVIRONMENTAL FLOWS MANAGEMENT PROBLEM + RESEARCH PROBLEM, to be completed by Chris / Lyndsey\>.

@@

The template is designed to be used within an adaptive preregistration methodology \<link to section\>.

In order to address the disjuncture between the iterative nature of model development, and the single-use, deterministic preregistration template, we take an 'expanded view of preregistration', that fits with the concept of 'adaptive preregistration' proposed by \<Srivastava\>. This view of preregistration breaks with the current model of preregistration, wherein the author writes a single deterministic preregistration containing a rule for every decision, and shifts from ideation, preregistration to execution of the plan. Two features define adaptive preregistration. It contains 'plans to deploy flexible strategies' \<Srivastava\> wherein the author can supply a heuristic consisting of multiple different analysis or modelling strategies whose execution depends on the outcome of previous decision-points or analyses. Secondly, the preregistration may be iterative and consist of interim preregistrations that mark phases of modelling and analysis as different parts of the data are observed \<Srivastava\>. Thus, as the modeller proceeds through the model development process, the researcher can shift from ideation, preregistration, execution of the analysis and back to ideation and preregistration again, depending on the observed outcomes as each interim preregistration is executed \<Link to Figure, GitHub as a platform for adaptive preregistration, <x-scrivener-item:///Users/egould/Dropbox/Apps/Scrivener/PhD_thesis.scriv?id=32FCFBDA-7CEE-4FD9-B8DD-F3073309C0D0>\>.

@@

Consequently we propose a methodology for implementing adaptive preregistration that leverages features of GitHub to also facilitate the task of linking the actual study and analysis conducted to the pre-specified plan in the preregistration \<link to part of GH methodology that explains how checking is accounted for in the procedural design\>. In addition, we designed a shiny interface for completing the preregistration to include a 'checking mode' that easily allows the author to state how and why the the analysis might have diverged from the planned \<link to part of shiny app interface that discusses checking mode\>.

@@

We propose leveraging both the *version-control* and the project management features of *GitHub* \<link to GitHub\> as a tool for implementing adaptive preregistration \<link to box\>. When a document is version-controlled using git and GitHub, each version of a document is time-stamped, and is assigned its own unique document identifier, or *commit-hash.* Because GitHub documents are 'time-stamped' the genesis of the preregistration from one version to the next is apparent. Moreover, using the *diff* view of a document on *GitHub*, exactly *how* a document has changed between versions is clear - additions are coloured green, deletions are coloured red, and changes within a line are highlighted \<insert picture\>. Since the preregistration document would exist within the project *repository* containing the data and modelling and analysis code itself, any results from analyses that cause the analysis plan to change can be linked and referenced in the updated version of the plan --- thus facilitating the process of model checking. By providing a method for contrasting the actual analysis undertaken against the registered analysis, this is a mechanism for allowing reviewers to properly evaluate  the preregistration against the reported analysis. It thus provides a way of marking preregistered parts of a report from non-preregistered parts of the analysis.

@@

Before shifting from planning to the execution of the interim preregistration, each complete response in the interim preregistration should have a corresponding *GitHub issue* created (Figure X, step 3, document). \<define GitHub issue, <https://guides.github.com/features/issues/>\>. Relevant discussion between analysts about how to proceed with the intended analysis or interpreting preliminary results are tracked within an *issue*'s thread. Each issue is assigned a unique number. Any code or analysis outputs, such as figures, tables or other files, should be committed and tagged with the issue number, so that all analysis addressing that task is tracked within the *issue's* thread (Fig X, step 4, Do) \<screen-cap?\>. The URL for the GitHub issue should then be added to the relevant preregistration item, the preregistration document should then be committed with the changes, and the version number of the preregistration updated in a new *release*.

@@

\- stating what needs to be known to move forward with the modelling and analysis

\- explaining how the researcher will test this, and what parts of the data will be used in that analysis

>>>>

---
title: "Preregistration for modelling in ecolgy and conservation"
description: |
  Working Draft.
author:
  - name: Elise Gould 
    affiliation: School of BioSciences, University of Melbourne
  - name: Chris Jones
    affiliation: Department of Environment Land and Water
  - name: Libby Rumpff
    affiliation: School of BioSciences, University of Melbourne
  - name: Hannah Fraser
    affiliation: School of BioSciences, University of Melbourne
  - name: Fiona Fidler
    affiliation: School of BioSciences, University of Melbourne
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    self_contained: false
---
**Intended journal:** Methods in Ecology and Evolution.

@@

A lack of complete and transparent reporting practices is concerning because it  provides the conditions in which QRPs are likely to arise [@Fidler:2017he]. Moreover, as Morrison [-@Morrison:2016cd] demonstrated, when models are not reported transparently with sufficient detail, they cannot be reproduced. This is particularly problematic because in ecology, models are typically used in decision-support, often directly underpinning policy and management decisions. Incomplete and non-transparent reporting of the model and the modelling process risk the model being used in inappropriate applications or decision contexts [@Grimm:2014es] because the underlying modelling choices and assumptions are not communicated [@deVos:2011]. Alternatively, a lack of transparency may mask poor model design or serious flaws in the model, with the risk that they be used in decision support resulting in wrong, or irreversible decisions [@Grimm:2014es;@Schmolke:2010fd]. A lack of transparency in reporting also risks undermining the credibility of the model, such that the decision-maker rejects the use of the model and instead turns to unstructured and subjective judgments [@Grimm:2014es;@Addison:2013wa], thus contributing further to "decision complacency", or the "knowing-doing" gap [@Gardner:2018dm].

Preregistration is a practice that has been adopted in disciplines that have confronted the 'reproducibility crisis' head-on, and can improve the transparency and reproducibility of ecological models. Preregistration requires the methods and analysis plan of a future study are registered in a secure and accessible platform prior to any data collection, and after submission it can no longer be altered. Preregistration distinguishes between what is a genuine a priori planned analysis, and what is not (Parker, Fraser, and Nakagawa 2019), and therefore prevents the opportunistic misuse of researcher degrees of freedom or 'Questionable Research Practices' (QRPs, Wicherts et al., 2016). Researcher degrees of freedom refer to the many analytic decisions a researcher makes during the planning, data collection, data analysis and reporting of results in scientific studies (Wicherts et al. 2016; Simmons, Nelson, and Simonsohn 2011). Examples of these analytic choices include whether to exclude observations, whether more data should be collected, which set of conditions should be combined and compared, which model evaluation metric should be used, etc. Coupled with a lack of transparency -- or "undisclosed flexibility" -- researcher degrees of freedom are vulnerable to being exploited by researchers because of cognitive biases and incentive structures within science. (lead in sentence to next para??)

Ecology has seen recent improvements in journal policy concerning reporting guidelines and data/code archiving, but there has been slow uptake of other important tools, like preregistration and registered reports (Parker, Fraser, and Nakagawa 2019). Pu et al. 2019 found that conceptual barriers may prevent users from different disciplines to think that preregistration is "not for them", this was a particularly salient sentiment for researchers engaged in non-experimental studies. We have experienced this sentiment ourselves when looking at preregistration templates to use for model-based research --- nearly all preregistration templates are geared towards experimental hypothesis-testing research. Such templates contain items that are simply irrelevant to model-based and/or applied ecological research. More importantly, they do not reflect the iterative nature of model development [@MacEachern2019] and fail to capture critical analytic decisions in the modelling process. Presently, we are aware of only a single preregistration template for modelling [@Cruwell2019b], however it is steeped heavily in an an academic psychology context, is focussed on the use of models for hypothesis-testing, neglects the8 model development process, and approaches modelling from a very narrow epistemic function, in a way that is not commonly applied in ecology.

Preregistration cannot simply be ported over to other disciplines and adopted as-is. If we wish to adopt preregistration in ecology, specifically within model-based research, both the content required to be preregistered, as well as the procedure for creating a preregistration must be adapted to suit model-based research as it is undertaken in ecology. Different research methodologies, and their domain-specific application across different domains and fields of study follow different analytical choices and are therefore susceptible to different QRPs. Consequently, domain-specific templates are necessary for guiding preregistration in such a way that restricts researcher degrees of freedom as they uniquely emerge within a domain or field-specific application of a particular research methodology.

@@

The goal of this study is to (rethink) preregistration to better suit the domain-specific challenges and methodological particularities of model-based research in ecology. This research builds on recently developed "QRP roadmaps" (Gould et al. 2019) and research by Cruwell et al. (2019) showing that existing templates are too general, and do not account for domain-specific and research category-specific nuances and questionable research practices. We aim to develop a template for preregistering ecological models, and devise a methodology for implementing an expanded view of the preregistration process, 'adaptive preregistration' (*sensu* Srivastava), that better accounts for the inherently iterative nature of model development than static, deterministic document-based preregistration.

@@

The secondary purpose of the model is to increase *research transparency* of model-based research in ecology and conservation. *Research transparency* refers to both production transparency --- which includes research artefacts like open-access data and materials, or data collection procedures --- and analytic transparency --- which describes a complete account of how the analytic conclusions are drawn from the data (Lupia and Elman, 2014). (refs to model reporting literature- it's dire, can't properly be evaluated by a reviewer, moreover can't be reproduced)

@@

More specifically, we intend this template to be suitable not just in 'pure-research'-only contexts, but to be suitable for ecologists and conservation decision-makers who are working in the 'actionable sciences' (which guy was this again.. Sutherland?? define).  The template should be adequate for use by scientists and modellers working in applied ecological contexts, but also for management and conservation decision-making contest where there are real-world decisions hinging on the model-based research.

(Research Types: Modelling - what do we mean by modelling?? I think if we don't specify this then people will just think 'classical linear regression' as modelling, and then will say, well why don't use existing hypothesis testing templates, like AsPredicted. I need to specify some particularities about the type of modelling that occurs in in these research contexts, and why existing templates don't suit, and that we need to design specifically for these research contexts.)

@@

Consequently, we aim not just to simply design a new template, deciding on its structure and content, but we propose a workflow and methodology to be implemented by the author when completing a preregistration of their model-based research (see section - inner logic of the model - discuss features model-based research, like iterative development?)

@@

**Specifying template objectives and performance criteria**

The overarching quality we wish the preregistration template to achieve is to adequately constrain researcher degrees of freedom during the model development process so as to reduce the the potential for questionable research practices and to transparently document the analytic decision process during modelling. Preregistration templates should be specific, precise and exhaustive [@Wicherts2016] if they are to adequately delimit researcher degrees of freedom. They should be *specific*, such that the preregistration is sufficiently detailed in describing all phases of the research process; *precise* such that each pre-specified aspect of the analysis plan is unambiguous and has only one interpretation; and *exhaustive* such that the preregistered analysis excludes the 'possibility of deviations from the preregistered research plan'. Although quantitative evidence about the effectiveness of preregistration is largely lacking, Bakker et al. [-@Bakker2018] quantitatively compared structured and unstructured preregistrations using a Transparency Score protocol that encapsulated these performance criteria, and concluded that detailed instructions and guidance within a template yields preregistrations that are better at restricting researcher degrees of freedom than unstructured formats with minimal guidance. Consequently we adopt these properties as are the means objectives for the template and aim to develop an *opinionated* template that provides detailed guidance and illustration such that preregistered analyses are sufficiently *specific,* *precise*, and *exhaustive* in order to curtail researcher degrees of freedom.  Preregistered analyses meeting these criteria should result in transparent and detailed records of any analysis decisions.

On the 4^th^ March 2020, we ran a series of structured activities designed to inform the development of two preregistration templates, one for field-work and sampling design, the other for modelling and data-analysis. We split the activities into two groups, to better focus discussion for each of these templates. All workshop materials and workshop outputs can be found on the OSF component here (LINK).

@@

The objective of this activity was to identify modelling workflows that capture critical decision-steps in the process of field sampling efforts and modelling exercises that should be included in a preregistration template. By "critical" we mean that without these decision-steps the template would not a) adequately constrain researcher degrees of freedom and b) transparently document the analytic decision-process. The activity consisted of two-parts, first, we individually reflected on our scientific process for a recent or memorable research project, listing the steps we undertook from the beginning to end.  In the second part of this exercise, we came together as a group to collate each of the steps in our personal workflows onto a group workflow template. This template was prepared earlier ((specific OSF link to the document)), and consisted of several key 'phases' of the scientific process, with decision-steps grouped under each phase. The development of the draft modelling workflow was derived from previous work by Gould et al. *in prep.,* whilst the field sampling workflow was created specifically for this exercise. We sorted and categorised each decision-step as belonging to a particular workflow phase. Next we grouped similar decision-steps that could be included as a single item on a preregistration template. At the end of this process we reviewed the suggested phases on the workflow to assess whether phases needed further discretisation, aggregation, removal, or amendment.

@@

This second activity aimed to identify critical challenges and roadblocks to the implementation and uptake of preregistration in ecology and conservation, as well as brainstorming solutions to these challenges. We conducted two lots of structured discussions, one for each of the field and modelling template groups. Each discussion was structured using a series of pre-prepared questions and talking points informed by recent debates about preregistration in the literature (cite / link to OSF, or even discuss??). These talking points were chosen with the aim of identifying problems in the content and or structuring of the phases and decision steps that were collectively identified in activity 1, as well as any conceptual or philosophical issues with preregistration in ecology, conservation, and in particular modelling, that might impact on the template structure, content and procedural implementation of preregistration.

@@

**1. Identify Scientific Workflows for ecological modelling, both real/in-practice and idealised.**

The goal of this analysis task was to generate a complete draft workflow for both fieldwork and modelling and data-analysis that will form the basis of the final preregistration templates. We used a combination of inductive and deductive coding (is this what it is called) to analyse the workshop outputs and generate a model scientific workflow for model-based research that would constitute the presumed scientific workflow that would be used to guide the structure of the preregistration template. We collated and coded the workshop outputs using (atlas.TI, insert software citation and version), (a qualitative coding software). All workshop outputs can be found here ((Link to Workshop Outputs OSF repository)).

A preliminary idealised scientific workflow for ecological modelling research was synthesised after reviewing the literature on 'good modelling practice' and structured decision-making literature in ecological management. Steps in the workflow were coded and broken down into 'phases' consisting of 'steps' and 'substeps', in line with the ontology we specified for translating a modelling workflow into a preregistration template [section link], (insert link to appendices or just present the final?). Each decision-point in the personal modelling workflows generated from activity one of the collaborative workshop were coded as belonging to one of these phases, steps and substeps. We revised the coding structure (inductively (is this the word)), according to patterns and themes identified across multiple personal workflows. For example, we added the final phase (name for last phase), which aimed to capture the fact that analysts usually present the results of their modelling to clients, decision-makers or other stakeholders that must interpret the evidence and are responsible for making management decisions. In this way, we identified a common or generalised workflow from different types of applied ecological modelling projects that synthesised both (idealised norms and norms of practice, Pu et al., see other philosophy of sci refs). This workflow is used to provide structure and content to the preregistration template.

**2. Translating Workflows into Preregistration Templates**

We translated the model-based workflow identified in part I of workshop analysis into a text-based preregistration template. Reflecting on existing templates (link to OSF repository of templates page), we identified common elements / units of preregistration templates that we mapped onto the generalised modelling workflow identified in part I of workshop output analysis. Following the features of a good guideline identified by Moallemi et al. [-@Moallemi2019], we provide examples of choices or illustrations to clarify each step of the preregistration, for parts of the modelling process that are semantically, epistemically and/or philosophically contentious or ambiguous (link to paper describing contentious issues in modelling history). We synthesised the following ontology to build our preregistration template:

1. Decision Phase - each has a "title" and "description" or "definition", steps are grouped under phases. (Note, that these are being scrapped from the template and are turned into tool-tips)

2. Decision Step - each has a "title" and "description" or "definition" choices are grouped into 'steps', which corresponds to an activity or set of related activities in a scientific workflow. (again, these are being scrapped and turned into tool-tips.)

3. Choices ---) each 'choice' corresponds to a uniquely numbered item on the PRT requiring a response from user. The style should be in the form of a directive or a question. E.g. "explain how you will do xyzzy". Or "What performance measure will you use for assessing 'goodness of fit'?"

**3. Designing a new methodology for implementing preregistration in model-based research**

@@

- What should a research workflow look like for a modeller in an applied ecology setting who wishes to preregister their work?

- How do we accommodate the iterative cycle of model development into the preregistration process?

- At what stage in a modelling or structured decision-making process should preregistration begin?

@@

(which other template design papers have included evaluation criteria and rationale, Sophia's?)

@@

Modellers [@MacEachern2019] (here, here, and here) have expressed concern about preregistration making modelling a more difficult task than it already is, and hampering the ability to meet project deadlines, especially in decision-problems. Many modellers already struggle with sufficiently documenting models during publication and reporting (ref, model reporting and transparency review), and modellers working in government or other agencies have expressed concern that preregistration will simply add to the list of  'red-tape' expected of them. Given these concerns, and considering the concerns of many with regard to the iterative nature of model development precluding making preregistration impossible or inappropriate, we figured it was of utmost importance that the templates be also evaluated on their ease-of-use and feasibility.

@@

We tested and evaluated the templates on the ability to meet these criteria by using a model-based research case-study. The user-centered design procedure we implemented follows two phases of evaluation. Firstly, the template is iteratively evaluated and refined 'in situ' -- throughout the process of preregistering the VEFMAP analysis with the template, any required changes to make the template more *specific*, *precise*, and *exhaustive* are implemented. After the preregistration is complete, the preregistration was subjected to a final round of quantitative evaluation in which we evaluated the ability of the template to meet its objectives using a Transparency Score protocol. We also subjectively reflected on the template and the process of preregistration.

*Case-study*

Chris / Lyndsey to complete -- 1 paragraph or 2 short paragraphs describing the decision problem / trigger and the modelling research question / problem.

*Iterative refinement*

Rather than 'digging back' through the version history of a completed preregistration to ask researchers about the process and how it had been developed in detail, we chose to leverage the version-control and collaborative project management features of Git and GitHub (Ben, Smith, and Bailey 2018; Blischak, Davenport, and Wilson 2016)  to 'live-develop' the template throughout the process of writing the preregistration for the case-study. Since the process of creating a preregistration is hard to reconstruct using *post hoc* interviews (Pu et al. 2019), we aimed to capture the analysts 'thinking in the moment' by using comments and GitHub issues describing why a particular decision to refine the template was made. For example, if it turned out that we had missed an important item, or perhaps the phrasing of an item or the order and structure of the template needed to change whilst in the process of completing the preregistration; the suggested change, and its justification was recorded in GitHub and simultaneously linked to the actual preregistration document change in the version-control record.

*Quantitative Evaluation using a Transparency Score Protocol*

We created a Transparency Score protocol following the method used by Bakker et al. [-@Bakker2018] to assess and compare the quality of preregistrations created with structured *versus* unstructured templates. We first created a 'QRP coding protocol' based on QRP roadmaps (Gould et al. *in prep.*), which we used to code the final preregistered analysis for the case-study (Table X). This process identified the presence and/or possibility of QRP's to occur within the preregistered analysis.

After coding for QRP's within the preregistration, we then quantitatively assessed the quality of the preregistration using a Transparency Score protocol (link to protocol in OSF). For each QRP in the QRP coding protocol, we assigned a score between 0 and 3 where:

* 0 = the QRP was not restricted at all
* 1 = the QRP was restricted to some degree
* 2 = the QRP was completely restricted (the template was both *specific* and *precise*)
* 3 = the QRP was completely restricted and the preregistration was *exhaustive*, i.e. it included explicit statements that no deviation from the way it was registered would occur.

For every QRP, these scores are qualified in specific detail to aid in scoring the template.

*Reflection*

@@

@@

The template is designed to be used within an adaptive preregistration methodology (link to section).

In order to address the disjuncture between the iterative nature of model development, and the single-use, deterministic preregistration template, we take an 'expanded view of preregistration', that fits with the concept of 'adaptive preregistration' proposed by (Srivastava). This view of preregistration breaks with the current model of preregistration, wherein the author writes a single deterministic preregistration containing a rule for every decision, and shifts from ideation, preregistration to execution of the plan. Two features define adaptive preregistration. It contains 'plans to deploy flexible strategies' (Srivastava) wherein the author can supply a heuristic consisting of multiple different analysis or modelling strategies whose execution depends on the outcome of previous decision-points or analyses. Secondly, the preregistration may be iterative and consist of interim preregistrations that mark phases of modelling and analysis as different parts of the data are observed (Srivastava). Thus, as the modeller proceeds through the model development process, the researcher can shift from ideation, preregistration, execution of the analysis and back to ideation and preregistration again, depending on the observed outcomes as each interim preregistration is executed (Link to Figure, GitHub as a platform for adaptive preregistration, <x-scrivener-item:///Users/egould/Dropbox/Apps/Scrivener/PhD_thesis.scriv?id=32FCFBDA-7CEE-4FD9-B8DD-F3073309C0D0>).

@@

Consequently we propose a methodology for implementing adaptive preregistration that leverages features of GitHub to also facilitate the task of linking the actual study and analysis conducted to the pre-specified plan in the preregistration (link to part of GH methodology that explains how checking is accounted for in the procedural design). In addition, we designed a shiny interface for completing the preregistration to include a 'checking mode' that easily allows the author to state how and why the the analysis might have diverged from the planned (link to part of shiny app interface that discusses checking mode).

@@

We propose leveraging both the *version-control* and the project management features of *GitHub* (link to GitHub) as a tool for implementing adaptive preregistration (link to box). When a document is version-controlled using git and GitHub, each version of a document is time-stamped, and is assigned its own unique document identifier, or *commit-hash.* Because GitHub documents are 'time-stamped' the genesis of the preregistration from one version to the next is apparent. Moreover, using the *diff* view of a document on *GitHub*, exactly *how* a document has changed between versions is clear - additions are coloured green, deletions are coloured red, and changes within a line are highlighted (insert picture). Since the preregistration document would exist within the project *repository* containing the data and modelling and analysis code itself, any results from analyses that cause the analysis plan to change can be linked and referenced in the updated version of the plan --- thus facilitating the process of model checking. By providing a method for contrasting the actual analysis undertaken against the registered analysis, this is a mechanism for allowing reviewers to properly evaluate  the preregistration against the reported analysis. It thus provides a way of marking preregistered parts of a report from non-preregistered parts of the analysis.

@@

Before shifting from planning to the execution of the interim preregistration, each complete response in the interim preregistration should have a corresponding *GitHub issue* created (Figure X, step 3, document). (define GitHub issue, <https://guides.github.com/features/issues/>). Relevant discussion between analysts about how to proceed with the intended analysis or interpreting preliminary results are tracked within an *issue*'s thread. Each issue is assigned a unique number. Any code or analysis outputs, such as figures, tables or other files, should be committed and tagged with the issue number, so that all analysis addressing that task is tracked within the *issue's* thread (Fig X, step 4, Do) (screen-cap?). The URL for the GitHub issue should then be added to the relevant preregistration item, the preregistration document should then be committed with the changes, and the version number of the preregistration updated in a new *release*.

@@

- stating what needs to be known to move forward with the modelling and analysis

- explaining how the researcher will test this, and what parts of the data will be used in that analysis
